Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-23 manuscript.rmd
1. added second para indentation 2. Removed spaces between set symbols such as i~|~z

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves. We validated ERD against existing tree and soil stable water isotope data. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

@@

# Introduction

\     
Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

@@

\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for 8 species and developing trait-based proxies for the rest. We evaluated ERD against existing data on plant and soil water isotopes and tested coordination of ERD with hydraulic traits and mortality rates. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:

@@

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour). The soil profile is discretized into 15 exponentially distributed soil layers with depth z $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).

### Model parameterization
\     To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ taht best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\
|   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model calibration
\
   We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).

@@

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) \\
& + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m})\right)
(\#eq:obj-func)
\end{split}
\end{equation}

based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s. $W_{liq,~\overline{z}}$ is the average of daily $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

\
We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.

## Model for effective rooting depth (ERD)
\
To estimate effective rooting depth, we tested alternate models for growth (Eqs. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance (K~leaf~) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth. In our models K~leaf~ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:

\begin{equation}
G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
(\#eq:M-chosen)
\end{equation}

where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term. See SI for equations for all the alternate models tested.

As VPD may have a non-linear relationship with growth, with growth increasing for the lower range of VPD, and decreasing for the higher range of VPD due to stomatal closure (reference), we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

For each species, we estimated diurnal maximum value of $K~{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi$~soil,~i~|~z,~h~ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

\begin{equation}
K_{leaf,~s,~i~|~z,~h} = A_{s}e^{-B_{s}\Psi_{soil,~i~|~z,~h}}
(\#eq:k-soil)
\end{equation}

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi$~soil,z~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

\begin{equation}
ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation} 
where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ 0.5, {`r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [AT:Chonggang, So that we can perform model evaluation, depths 0.21, 0.37 and 0.62 were all taken to be as 0.5 m (not averaged). How do I explain that? So hereafter 0.21, 0.37 and 0.62 were not retained, but only 0.5 m. How do I explain that? Or should I retain them?].

@@

\begin{equation}
K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the 31 ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

>>>>

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves. We validated ERD against existing tree and soil stable water isotope data. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008]. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

@@

\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile. We tested coordination of ERD with hydraulic traits for a subset of eight species with both datasets. We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W$~can~, surface water $\Delta W$~sfc~, soil water $\Delta W$~liq,z~ at depth z and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes). Conservation of these terms is expressed as follows:

@@

where, $q$~rain~ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q$~over~ is surface runoff, $q$~h2osfc~ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour). The soil profile is discretized into 15 exponentially distributed soil layers with depth z, where z $\in$ Z, and Z = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model calibration
\
We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq. \@ref(eq:water-balance)), (ii) stream discharge $q_{run} from the Conrad catchment, comprabel to $q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}) comparable to W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to plot-wide averages).

@@

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,E_{v} + E_{g}) + RMSE^{*}(q_{run},q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,\overline{z}}}) \\
& + RMSE^{*}(\theta_{0.15m}, W_{liq,`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m})\right)
(\#eq:obj-func)
\end{split}
\end{equation}

based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s. $W$~liq,~$_{\overline{z}}$ is the average of daily $W$~liq,z~, where z $\in$ Z, and Z = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

\   Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m}) < 0.2$), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

## Hydrological droughts
\   We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth *z* as days that fell below the 5 percentile of mean $\Psi$~soil,*z*~ across all hydrological realisations values at that depth.

## Model for effective rooting depth (ERD)
\
To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance (K~leaf~) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth. In our models K~leaf~ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species', growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

\    For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} VPD_{i}^{*}) + \epsilon_{t|z}
(\#eq:M-chosen)
\end{equation}

where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term. See SI for equations for all the alternate models tested.

\    As VPD may have a non-linear relationship with growth, with growth increasing with VPD intially, but decreasing with VPD beyond an optimal value due to stomatal closure [AT:Chonggang, add a reference], we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

\    For each species, we estimated diurnal maximum value of K~leaf~$_{,s,i|z,h}$ for a species $s$ by substituting $\Psi$~soil~$_{,s,i|z,h}$ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

\begin{equation}
K_{leaf,s,i|z,h} = A_{s}e^{-B_{s}\Psi_{soil,i|z,h}}
(\#eq:k-soil)
\end{equation}

\    By varying *z*, we estimated species effective rooting depths (ERD), defined as the depth *z* at which soil water dynamics ($\Psi$~soil,*z*~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species *s*, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient *r* for each soil depth *z* for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood (the minimum AIC score), and within those depths retained the ones that had Pearson's *r* >= 0.71 (or, *R*^2 >= 0.5) as a valid set. For a species *s*, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

\begin{equation}
ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s|z,h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation} 
where $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.5, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [AT:Chonggang, so that we can perform model evaluation, depths 0.21, 0.37 and 0.62 were all taken as if they were 0.5 m (not averaged).  How do I explain that? So hereafter 0.21, 0.37 and 0.62 were not retained, but only 0.5 m.].

@@

\begin{equation}
K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

\   As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the 31 ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

\   We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq. \@ref(eq:M-chosen).
